Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-19 manuscript.qmd
fix references

When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted. Per current guidelines, large volumes of the neck - lymph node levels (LNLs) - are part of the the elective clinical target volume (CTV-N) [@gregoire_ctbased_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensitymodulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009]. This is to minimize the risk of regional recurrences due to missed microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).

@@

The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multicentric_2023. Although these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.

@@

It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity. However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference. We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging. This is virtually impossible with sensitivities around 80% and specificities lower than 100%.

Our pre-training consensus was formed by considering all reported diagnostic information for a particular patient and LNL. When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].

@@

For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013]. It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs. The sample proposal algirothms used by us are based on differential evolution moves [@terbraak_differential_2008;@nelson_run_2013]. The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.

@@

@Fig-model-risks shows this predicted risk of occult disease in contralateral LNL II for four interesting combinations of these three risk factors. This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76% [@debondt_detection_2007].

>>>>

When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted. Per current guidelines, large volumes of the neck - lymph node levels (LNLs) - are part of the the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009]. This is to minimize the risk of regional recurrences due to missed microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET).

@@

The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023. Although these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.

@@

It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity. However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference. We opted for this because the literature values of sensitivity and specificity [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging. This is virtually impossible with sensitivities around 80% and specificities lower than 100%.

Our pre-training consensus was formed by considering all reported diagnostic information for a particular patient and LNL. When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].

@@

For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013]. It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs. The sample proposal algirothms used by us are based on differential evolution moves [@ter_braak_differential_2008;@nelson_run_2013]. The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.

@@

@Fig-model-risks shows this predicted risk of occult disease in contralateral LNL II for four interesting combinations of these three risk factors. This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].
